One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2015-01-13 paper/model.tex
still revising...

%isolated system 
% ang mom cons
% The locally isothermal
% equation of state allows us to apply standard linear density wave
% theory for results interpretation. 
% This allows us to consider
%an isolated disc.

@@

\section{Linear density waves}\label{wkb}
% In this paper we will mainly use non-linear hydrodynamic simulations to investigate the
% stability of disc equilibria described 
% above. However, it is useful to  interpret simulation results
% using linear density theory. We list here the main definitions and
% results from 2D linear theory, taken from \cite{lin11b} and
% \cite{shu91}.  
{\bf We highlight a key feature of locally isothermal discs that
  enables angular momentum exchange between small disturbances and the
  background disc through an imposed radial temperature gradient. This
  conclusion results from the consideration of angular momentum
  conservation within the framework of linear perturbation theory. 
  For simplicity, in this section we consider 2D discs. 
}

In a linear analysis, one assumes a steady axisymmetric background state, % such as
% that defined in \S\ref{model},
which is then perturbed such that
\begin{align}  
  \Sigma \to \Sigma(R) + \delta\Sigma_m(R)\exp{\left[\ii\left(-\sigma t +
        m\phi\right)\right]}, 
\end{align}
and similarly for other variables, where $\sigma=\omega+\ii\gamma$ is
a complex frequency with $\omega$ being the real frequency, 
$\gamma$ the growth rate, and $m$ is an integer. We take $m>0$ without
loss of generality.  
%Since the background is
%axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
 %   t\right)} = \Sigma_m$ as defined for diagnostics.

@@

These linearised equations can be combined into a single
integro-differential equation eigenvalue problem. We defer a full
{\bf numerical} exploration of the linear problem to a future study. Here, we {\bf
  discuss} some
general properties of the linear perturbations. % that will be relevant
% for describing simulation results.

@@

In Eq. \ref{angcons}, the background torque density $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\delta\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state. {\bf  We outline the
derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.}

@@

The important consequence of the background torque is the possibility
of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
(negative) and the background torque $\tbg$ is {\bf also} positive (negative),
then the local angular momentum density of the linear disturbance
{\bf will} further increase (decrease) {\bf with time}. This implies the amplitude of the disturbance
may grow by exchanging angular momentum with the background
disc.%  In our disc models with $c_s^2\propto R^{-q}$, we expect the
% growth rate of such an instability to scale with $q$.

{\bf 
  We demonstrate instability for low-frequency modes 
  ($|\omega|\ll m\Omega$) by explicitly showing its 
  angular momentum density 
  $\jlin<0$, and the background torque $T_\mathrm{BG}<0$ for
  appropriate perturbations and radial temperature gradients. 
%, and show that
%  low-frequency modes can be unstable when a radial 
%  temperature gradient is imposed.  
}

@@

Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv  
\omega/m$ of a neutral mode 
, Eq. \ref{dispersion} can be solved 
for $|k|$, 
\begin{align}\label{wavenumber}
%  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
 %     Q^2(1-\nu^2)}\right], 
|k| = k_c\left[1 \pm \sqrt{1 -
     Q^2(1-\nu^2)}\right], 
\end{align}
where 
 \begin{align}
   k_c \equiv \frac{\pi G \Sigma}{c_s^2}
 \end{align}
 is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. In
Eq. \ref{wavenumber}, the upper (lower) sign correspond to short
(long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
waves.

The \emph{co-rotation radius} $R_c$ is where the pattern speed matches
the fluid rotation,
\begin{align}
  \Omega(R_c) = \Omega_p.
\end{align}
\emph{Lindblad resonances} $R_L$ occurs where
\begin{align}
  \nu^2(R_L) = 1. 
\end{align}
Finally, \emph{Q-barriers} occur at radii $R_{Qb}$ where
\begin{align}
  Q^2(R_{Qb})\left[1-\nu^2(R_{Qb})\right] = 1.  
\end{align}
According to Eq. \ref{wavenumber}, purely wave-like solutions with
real $k$ are only possible where $Q^2(1-\nu^2)\leq1$.

A detailed discussion of the properties of local density waves 
is given in \cite{shu91}. An important result, {\bf which holds for
  waves of all frequencies}, is that
waves interior to co-rotation ($R<R_c$) have negative angular momentum, while
waves outside co-rotation ($R>R_c$) have positive angular
momentum. 
%The local approximation becomes less good for global modes,
%but will nevertheless be useful for results interpretation.

\subsection{Unstable interaction between low-frequency modes 
  and the background disc due to an imposed temperature gradient}
%The $m=1$ spiral that emerges in our simulations have low frequency
%($\Omega_p < \Omega$), grows slowly ($\gamma\ll\Omega_p$) and its
%spatial form is consistent local theory (\S\ref{fargo_m1}). 
%{\bf
%  We first regard a low-frequency, $m=1$ disturbance 
 % as a mode within the framework of local theory, satisfying the
 % % temperature gradient ignored in local theory
 % dispersion relation Eq. \ref{dispersion}. }
%To demonstrate possible
%instability, 
{\bf 
Here we show that the torque density acting on a low-frequency, 
local mode due to the background temperature gradient can be negative, which 
would enforce the mode, because it has negative angular momentum.  
}

% Note that
% we will not be assuming a self-gravitating disc until the end of this
% section. 
%We begin by recalling the expression for the torque density due to a
%locally isothermal equation of state,
%\begin{align*}
 % T_\mathrm{BG} = - \frac{m}{2}\frac{dc_s^2}{dR}\imag\left(\delta\Sigma_m\xi_R^*\right).   
%\end{align*}
The Eulerian surface density perturbation is given by
\begin{align}\label{den_pert}
  \delta\Sigma_m = -\nabla\cdot\left(\Sigma\bm{\xi}\right) 
  = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma \xi_R\right) - \frac{\ii m}{R}\Sigma\xi_\phi.  
\end{align}
We invoke local theory by setting $d/dR \to \ii k$ where $k$ is
real, and assume $|kR|\gg m$ so that the second term on the right hand
side of Eq. \ref{den_pert} can be neglected. Then    
\begin{align}
  \delta\Sigma _m  \simeq -\ii k \Sigma \xi_R.
\end{align}
{\bf Inserting this into the expression for the background torque,
  Eq. \ref{baroclinic_torque}, we find}
\begin{align}
  T_\mathrm{BG} = \frac{m}{2}\frac{dc_s^2}{dR}k\Sigma |\xi_R|^2. \label{baroclinic_torque1}
\end{align}

@@

Using $\jlin$ and $T_\mathrm{BG}$, we can estimate the 
growth rate $\gamma$ of linear perturbations due to the background
torque as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadratic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for gives
\begin{align}\label{theoretical_rate0}
  2\gamma \sim
  -\frac{dc_s^2}{dR}
  \frac{k}{\Omega}\frac{|\xi_R|^2}{\left[(m-1)|\bm{\xi}|^2 + |\xi_R +
      \ii\xi_\phi|^2 \right]} 
  = -\frac{dc_s^2}{dR}
  \frac{k}{m\Omega},
%  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
%   \xi_\phi|^2}. 
\end{align}
{\bf where the second equality uses $\xi_\phi \simeq
2\ii \xi_R/m$ in the local approximation, as shown in  Appendix
\ref{horizontal_displacements}.} 
 % {\bf
% This holds for local modes of all frequencies and can be shown
% explicitly by evaluating Eq. \ref{ang_mom_def} with 
% \begin{align}
%   \xi_\phi = -\frac{2\ii\Omega}{\sbar}\xi_R, 
% \end{align}
% consistent with the local approximation as shown in Appendix
% \ref{horizontal_displacements}. 
% }
Then 
%To further simplify this expression, we need to relate $\xi_R$ and
%$\xi_\phi$. 
% \begin{align}
%   2\gamma \sim
%   -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
%     + \ii\xi_\phi|^2}.
% \end{align}
% Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
% in the local approximation, we have
% \begin{align}
%   2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
%     \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
%   \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
% \end{align}
% where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
% inequality. 
%where we used $dc_s^2/dR <0$ and $k>0$ to obtain the inequality. 
for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
adopted in our disc models,  
\begin{align}
  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{m\Omega} \sim q h
  \left(\frac{kH}{m}\right)\Omega,\label{theoretical_rate}
\end{align}
where we used $c_s\sim H\Omega\sim hR\Omega$. {\bf
  Eq. \ref{theoretical_rate} suggests that perturbations with small
  radial length-scales ($kH\gtrsim m$) are most favourable for
  destabilisation. Taking the local approximation is then
  appropriate.}

@@

% For perturbations in a self-gravitating disc with characteristic
% wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$, as seen in
% our simulations, we have 
% \begin{align}\label{theoretical_rate}
%   \gamma \sim \frac{qh}{2Q}\Omega,
% \end{align}
% where we used $c_s/R\sim h\Omega$.

%For our fiducial disc models with
%$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega \sim O(10^{-2})$,
%consistent with simulation results. Specifically,
%Eq. \ref{theoretical_rate} indicates the growth rate should increase
%linearly with $q$ and $h$, as seen in the simulations
%(Fig. \ref{fargo_varq_growth}---\ref{fargo_varh_growth}).

\section{Numerical simulations}\label{methods}
%above discussion is generic for low-frequency non-axisymmetric modes,
%but illustrate with a concrete example
{\bf
  We illustrate the destabilising effect of a fixed temperature
  gradient using numerical simulations. The above discussion is
  generic for low-frequency non-axisymmetric modes, but for
  simulations we will consider specific examples.

  There are two components to the destabilising mechanism.  First
  the disc should support low-frequency modes, which is then
  destabilised by an imposed temperature gradient. The latter is
  straight forward to implement by adopting a locally isothermal
  equation of state as described in \S\ref{model}.  For the former, we
  consider discs with a  
  radially-structured Toomre $Q$ profile. We use local theory to show that such discs support
  low-frequency trapped one-armed ($m=1$) modes. 
  This is convenient because Eq. \ref{theoretical_rate} indicates that
  modes with small $m$ are
  most favourable for destabilisation by an imposed temperature
  gradient.  
}

@@

%In 3D, we assume vertical hydrostatic balance
The 3D disc structure is obtained by assuming vertical hydrostatic
balance  
\begin{align}
  0 = \frac{1}{\rho}\frac{\p p}{\p z} + \frac{\p\Phi_*}{\p z} + \frac{\p
    \Phi_d}{\p z},  
\end{align}
which gives the mass density as 
\begin{align}
  \rho = \frac{\Sigma}{\sqrt{2\pi}H}Z(R,z),
\end{align}
where $Z(R,z)$ describes vertical stratification. In practice, we
numerically solve for $Z(R,z)$ by neglecting the radial self-gravity
force compared to vertical self-gravity, which reduces the equations
for vertical hydrostatic equilibrium to ordinary differential
equations. This procedure is described in \cite{lin12b}.

@@

\subsection{Codes}
{\bf We use three independent grid-based codes to simulate the aforementioned
system.} We adopt computational units such that
$G=M_*=R_0=1$. Time is measured  in the Keplerian orbital period at
the reference radius, $P_0\equiv  2\pi/\Omega_k(R_0)$.

>>>>

@@

\section{Linear density waves}\label{wkb}

{\bf We describe a key feature of locally isothermal discs that
  enables angular momentum exchange between small disturbances and the
  background disc through an imposed radial temperature gradient. This
  conclusion results from the consideration of angular momentum
  conservation within the framework of linear perturbation theory. 
  For simplicity, in this section we consider 2D discs. 
}

In a linear analysis, one assumes a steady axisymmetric background state, 
which is then perturbed such that
\begin{align}  
  \Sigma \to \Sigma(R) + \delta\Sigma_m(R)\exp{\left[\ii\left(-\sigma t +
        m\phi\right)\right]}, 
\end{align}
and similarly for other variables, where $\sigma=\omega+\ii\gamma$ is
a complex frequency with $\omega$ being the real frequency, 
$\gamma$ the growth rate, and $m$ is an integer. We take $m>0$ without
loss of generality.

@@

These linearised equations can be combined into a single
integro-differential equation eigenvalue problem. We defer a full
{\bf numerical} exploration of the linear problem to a future study. Here, we {\bf
  discuss} some
general properties of the linear perturbations.

@@

In Eq. \ref{angcons}, the background torque density $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\delta\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state {\bf in the perturbed disk}. {\bf  We outline the
derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.}

@@

The important consequence of the background torque is the possibility
of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
(negative) and the background torque $\tbg$ is {\bf also} positive (negative),
then the local angular momentum density of the linear disturbance
{\bf will} further increase (decrease) {\bf with time}. This implies the amplitude of the disturbance
may grow by exchanging angular momentum with the background
disc.

{\bf 
  We demonstrate instability for low-frequency modes 
  ($|\omega|\ll m\Omega$) by explicitly showing its 
  angular momentum density 
  $\jlin<0$, and the background torque $T_\mathrm{BG}<0$ for
  appropriate perturbations and radial temperature gradients. 
}

@@

Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv  
\omega/m$ of a {\bf non-axisymmetric} neutral mode 
, Eq. \ref{dispersion} can be solved 
for $|k|$, 
\begin{align}\label{wavenumber}
%  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
 %     Q^2(1-\nu^2)}\right], 
|k| = k_c\left[1 \pm \sqrt{1 -
     Q^2(1-\nu^2)}\right], 
\end{align}
where 
 \begin{align}
   k_c \equiv \frac{\pi G \Sigma}{c_s^2}
 \end{align}
 is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. In
Eq. \ref{wavenumber}, the upper (lower) sign correspond to short
(long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
waves.

{\bf At the \emph{co-rotation radius} $R_c$} the pattern speed matches
the fluid rotation,
\begin{align}
  \Omega(R_c) = \Omega_p.
\end{align}
\emph{Lindblad resonances} $R_L$ occurs where
\begin{align}
  \nu^2(R_L) = 1. 
\end{align}
Finally, \emph{Q-barriers} occur at radii $R_{Qb}$ where
\begin{align}
  Q^2(R_{Qb})\left[1-\nu^2(R_{Qb})\right] = 1.  
\end{align}
According to Eq. \ref{wavenumber}, purely wave-like solutions with
real $k$ are only possible where $Q^2(1-\nu^2)\leq1$.

A detailed discussion of the properties of local density waves 
is given in \cite{shu91}. An important result, {\bf which holds for
  waves of all frequencies}, is that
waves interior to co-rotation ($R<R_c$) have negative angular momentum, while
waves outside co-rotation ($R>R_c$) have positive angular
momentum.

\subsection{Unstable interaction between low-frequency modes 
  and the background disc due to an imposed temperature gradient}
{\bf 
Here we show that the torque density acting on a low-frequency, 
local mode due to the background temperature gradient can be negative, which 
would enforce the mode, because it has negative angular momentum.  
}

The Eulerian surface density perturbation is given by
\begin{align}\label{den_pert}
  \delta\Sigma_m = -\nabla\cdot\left(\Sigma\bm{\xi}\right) 
  = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma \xi_R\right) - \frac{\ii m}{R}\Sigma\xi_\phi.  
\end{align}
We invoke local theory by setting $d/dR \to \ii k$ where $k$ is
real, and assume $|kR|\gg m$ so that the second term on the right hand
side of Eq. \ref{den_pert} can be neglected. Then    
\begin{align}
  \delta\Sigma _m  \simeq -\ii k \Sigma \xi_R.
\end{align}
{\bf Inserting this into the expression for the background torque,
  Eq. \ref{baroclinic_torque}, we find}
\begin{align}
  T_\mathrm{BG} = \frac{m}{2}\frac{dc_s^2}{dR}k\Sigma |\xi_R|^2. \label{baroclinic_torque1}
\end{align}

@@

Using $\jlin$ and $T_\mathrm{BG}$, we can estimate the 
growth rate $\gamma$ of linear perturbations due to the background
torque as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadratic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for gives
\begin{align}\label{theoretical_rate0}
  2\gamma \sim
  -\frac{dc_s^2}{dR}
  \frac{k}{\Omega}\frac{|\xi_R|^2}{\left[(m-1)|\bm{\xi}|^2 + |\xi_R +
      \ii\xi_\phi|^2 \right]} 
  = -\frac{dc_s^2}{dR}
  \frac{k}{m\Omega},
%  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
%   \xi_\phi|^2}. 
\end{align}
{\bf where the second equality uses $\xi_\phi \simeq
2\ii \xi_R/m$ in the local approximation, as shown in  Appendix
\ref{horizontal_displacements}.} 
Then for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
adopted in our disc models,  
\begin{align}
  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{m\Omega} \sim q h
  \left(\frac{kH}{m}\right)\Omega,\label{theoretical_rate}
\end{align}
where we used $c_s\sim H\Omega\sim hR\Omega$. {\bf
  Eq. \ref{theoretical_rate} suggests that perturbations with small
  radial length-scales ($kH\gtrsim m$) are most favourable for
  destabilisation. Taking the local approximation is then
  appropriate.}

@@

\section{Numerical simulations}\label{methods}

{\bf
  We demonstrate the destabilising effect of a fixed temperature
  gradient using numerical simulations. The above discussion is
  generic for low-frequency non-axisymmetric modes, but for
  simulations we will consider specific examples.

  There are two components to the destabilising mechanism.  First
  the disc should support low-frequency modes, which is then
  destabilised by an imposed temperature gradient. The latter is
  straight forward to implement by adopting a locally isothermal
  equation of state as described in \S\ref{model}.  For the former, we
  consider discs with a  
  radially-structured Toomre $Q$ profile. We use local theory to show that such discs support
  low-frequency trapped one-armed ($m=1$) modes. 
  This is convenient because Eq. \ref{theoretical_rate} indicates that
  modes with small $m$ are
  more favourable for destabilisation by an imposed temperature
  gradient.  
}

@@

The 3D disc structure is obtained by assuming vertical hydrostatic
balance  
\begin{align}
  0 = \frac{1}{\rho}\frac{\p p}{\p z} + \frac{\p\Phi_*}{\p z} + \frac{\p
    \Phi_d}{\p z},  
\end{align}
which gives the mass density as 
\begin{align}
  \rho = \frac{\Sigma}{\sqrt{2\pi}H}Z(R,z),
\end{align}
where $Z(R,z)$ describes vertical stratification. In practice, we
numerically solve for $Z(R,z)$ by neglecting the radial self-gravity
force compared to vertical self-gravity, which reduces the equations
for vertical hydrostatic equilibrium to ordinary differential
equations. This procedure is described in \cite{lin12b}.

@@

\subsection{Codes}
{\bf We use three independent grid-based codes to simulate the above 
system.} We adopt computational units such that
$G=M_*=R_0=1$. Time is measured  in the Keplerian orbital period at
the reference radius, $P_0\equiv  2\pi/\Omega_k(R_0)$.

### 2015-01-13 paper/paper.tex
still revising...

\maketitle
\begin{abstract} 
  {\bf We describe a new mechanism that leads to the destabilisation of
 non-axisymmetric waves in astrophysical discs with 
 an imposed radial temperature gradient. This might apply, for
 example, to the outer parts of protoplanetary discs. We use linear
 density wave theory 
 to show that non-axisymmetric perturbations generally do not conserve
 their angular momentum in the presence of a forced temperature
 gradient. This implies an exchange of angular momentum
 between linear perturbations and the background disc. In particular, 
 when the disturbance is a low-frequency
 trailing wave and the disc temperature decreases outwards, this 
 interaction is unstable and leads to the growth of the wave.  
 We demonstrate this phenomenon through numerical hydrodynamic
 simulations of locally isothermal discs in 2D using the FARGO
 code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
 radially structured discs with a self-gravitating region which remains
 stable in the absence of a temperature gradient. However,
 when a temperature gradient is imposed we observe exponential growth of an one-armed  
 spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
 outside the bulk of the spiral arm, resulting in a nearly-stationary
 one-armed spiral pattern.  The development of this one-armed 
 spiral does not require the movement of the central star, as found in previous studies. 
 Because destabilisation by the imposed temperature gradient does 
 not explicitly require disc self-gravity, we suggest this mechanism may also
 affect low-frequency one-armed oscillations in non-self-gravitating
 discs. 
}
\end{abstract}

%This might represent, for example, mass  
 % built-up in a circumstellar or protoplanetary
 % disc at the outer edge of a `dead zone', in which there
  %is inefficient magneto-hydrodynamic turbulence to enable accretion.

@@

\section{Introduction}\label{intro}
% Protoplanetary discs are expected to contain complex structures
% \citep{armitage11}, which directly impact planet formation theories.  
An exciting development in the study of circumstellar 
discs is the direct observation of large-scale, non-axisymmetric
structures within them. These
include lopsided dust distributions 
\citep{marel13,fukagawa13,casassus13,isella13,perez14,follette14} and
spiral arms 
\citep{hashimoto11,muto12,boccaletti14,grady13,christiaens14,avenhaus14}.

@@

Single-arm spirals, or eccentric modes, corresponding to perturbations 
with azimuthal wavenumber $m=1$, have received interest in
the context of protoplanetary discs because of their global nature
\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. 
In the `SLING' mechanism proposed by \cite{shu90}, the $m=1$ {\bf
  gravitational spiral} instability is attributed to motion of the
central star induced by the one-armed perturbation and requires a
massive disc \citep[the former may have observable consequences,
][]{michael10}.

% The aforementioned analytical studies have adopted
% barotropic, smooth disc models.
% In this work, we identify a mechanism to generate one-
% armed spiral patterns that is related to a non-barotropic
% equation of state and disc structure.
% We find that if the disc temperature or sound-speed is a
% prescribed function of position, then one-armed spiral modes
% can grow inside a cavity that results from the adopted disc
% structure.

%The aforementioned analytical studies have adopted barotropic, smooth
%disc models.

{\bf
  In this work we identify a new mechanism that leads to the growth of
  one-armed spirals in astrophysical discs. We show
  that when the disc temperature is prescribed (called locally isothermal
  discs), the usual statement for the conservation of angular momentum
  for linear perturbations acquires a source term proportional to the
  temperature gradient. This permits angular momenum exchange between
  linear perturbations and the background disc. This `background
  torque' may act to destabilise low-frequency non-axisymmetric
  trailing waves when the disc temperature decrease outwards.       
}

{\bf We employ direct hydrodynamic simulations using three different
  grid-based codes to demonstrate how this `background torque' 
  can lead to the growth of one-armed spirals in radially structured, 
  self-gravitating discs. This is despite the fact that our disc
  models do not meet the requirements for the `SLING'
  mechanism. Although our numerical simulations consider
  self-gravitating discs, this `background torque' is generic for
  locally isothermal discs and its existence does not require
  disc self-gravity. Thus, the destabilisation effect we
  describe should also be applicable to non-self-gravitating discs.  
}

@@

% In the case of protoplanetary discs, which have complex structure
% \citep{armitage10}, GI can also occur in `dead
% zones' --- sub-portions of the disc where the magneto-rotational
% instability is ineffective for angular momentum transport
% \citep{gammie96,turner08,landry13} and therefore has a reduced mass
% accretion rate. Dead zones may then become
% self-gravitating with sufficient mass built-up
% \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.

% original motivation for this work (DZGI)

%A wide range of GIs can occur in  gaseous 
%astrophysical discs 
%\citep{papaloizou89,christo92,christo93,hadley11,hadley14}.

% It is then
% of theoretical importance to clarify what type of GI is relevant to 
% partially self-gravitating discs, such as a protoplanetary disc containing a 
% massive dead zone. 
% This work presents a first step towards this goal. Specifically, here we 
% study a class of seemingly simple disc models which were found to 
% consistently produce \emph{one-armed} spiral structures, despite not 
% satisfying the usual requirements to generate them.

%through this work we identify a new method to generate one arm
%spirals 
%We therefore dedicate this study to verify their origin.

%brief history of m=1 
%stellar wobble from indiana group

>>>>

\maketitle
\begin{abstract} 
  {\bf We describe a new mechanism that leads to the destabilisation of
 non-axisymmetric waves in astrophysical discs with 
 an imposed radial temperature gradient. This might apply, for
 example, to the outer parts of protoplanetary discs. We use linear
 density wave theory 
 to show that non-axisymmetric perturbations generally do not conserve
 their angular momentum in the presence of a forced temperature
 gradient. This implies an exchange of angular momentum 
 between linear perturbations and the background disc. In particular, 
 when the disturbance is a low-frequency 
 trailing wave and the disc temperature decreases outwards, this 
 interaction is unstable and leads to the growth of the wave.  
 We demonstrate this phenomenon through numerical hydrodynamic
 simulations of locally isothermal discs in 2D using the FARGO
 code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
 radially structured discs with a self-gravitating region which remains
 stable in the absence of a temperature gradient. However,
 when a temperature gradient is imposed we observe exponential growth of a one-armed  
 spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
 outside the bulk of the spiral arm, resulting in a nearly-stationary
 one-armed spiral pattern.  The development of this one-armed 
 spiral does not require the movement of the central star, as found in previous studies. 
 Because destabilisation by a forced temperature gradient does 
 not explicitly require disc self-gravity, we suggest this mechanism may also
 affect low-frequency one-armed oscillations in non-self-gravitating
 discs. 
}
\end{abstract}

@@

\section{Introduction}\label{intro}
An exciting development in the study of circumstellar 
discs is the direct observation of large-scale, non-axisymmetric
structures within them. These
include lopsided dust distributions 
\citep{marel13,fukagawa13,casassus13,isella13,perez14,follette14} and
spiral arms 
\citep{hashimoto11,muto12,boccaletti14,grady13,christiaens14,avenhaus14}.

@@

Single-arm spirals, or eccentric modes, corresponding to perturbations 
with azimuthal wavenumber $m=1$, have received interest in 
the context of protoplanetary discs because of their global nature
\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. 
In the `SLING' mechanism proposed by \cite{shu90}, a $m=1$ {\bf
  gravitational} instability arises from the motion of the
central star induced by the one-armed perturbation, and requires a
massive disc \citep[the former may have observable consequences,
][]{michael10}.

{\bf
  In this work we identify a new mechanism that leads to the growth of
  one-armed spirals in astrophysical discs. We show
  that when the disc temperature is prescribed (called locally isothermal
  discs), the usual statement of the conservation of angular momentum
  for linear perturbations acquires a source term proportional to the
  temperature gradient. This permits angular momenum exchange between
  linear perturbations and the background disc. This `background
  torque' may act to destabilise low-frequency non-axisymmetric
  trailing waves when the disc temperature decreases outwards.       
}

{\bf We employ direct hydrodynamic simulations using three different
  grid-based codes to demonstrate how this `background torque' 
  can lead to the growth of one-armed spirals in radially structured, 
  self-gravitating discs. This is despite the fact that our disc
  models do not meet the requirements for the `SLING' 
  mechanism. Although our numerical simulations consider
  self-gravitating discs, the `background torque' is generic for
  locally isothermal discs and its existence does not require
  disc self-gravity. Thus, the destabilisation effect we
  describe should also be applicable to non-self-gravitating discs.  
}

@@
